---
title: "2026 Georgia Lottery Sales Forecast"
date: "April 14, 2026"
format:
  html:
    toc: false
    theme: cosmo
    self-contained: true
execute:
  echo: false
---

```{r}
#| label: load-packages
#| message: false
#| warning: false
library(tidyverse)
library(leaflet)
library(sf)
library(tigris)
```

```{r}
#| label: load-data
#| message: false
#| warning: false
georgia_lottery <- readRDS("georgia_lottery.RDS")
source("create_lottery_map.R")
```

```{r}
#| label: build-model
#| message: false
#| warning: false

# Aggregate to year + demographic group level
annual_totals <- georgia_lottery |>
  group_by(county, year, sex, race_ethnicity, age_group, income_level) |>
  summarise(
    lottery_spend = sum(lottery_spend, na.rm = TRUE),
    n = sum(n, na.rm = TRUE),
    .groups = "drop"
  )

# Fit a single multiple linear regression on the full dataset
fit <- lm(lottery_spend ~ year + sex + race_ethnicity + age_group + income_level + county,
          data = annual_totals)
n_fit <- lm(n ~ year + sex + race_ethnicity + age_group + income_level + county,
            data = annual_totals)

# Build a prediction grid for 2026 covering all unique strata
pred_grid <- annual_totals |>
  distinct(county, sex, race_ethnicity, age_group, income_level) |>
  mutate(year = 2026L)

predictions_2026 <- pred_grid |>
  mutate(
    lottery_spend = pmax(predict(fit, newdata = pred_grid), 0),
    n             = pmax(predict(n_fit, newdata = pred_grid), 0)
  )

total_predicted <- sum(predictions_2026$lottery_spend, na.rm = TRUE)
```

Based on historical trends, Georgia residents are projected to spend **\$`{r} format(round(total_predicted / 1e6, 1), nsmall = 1)` million** on lottery tickets in 2026.

```{r}
#| label: scatter-trend
#| fig-cap: "Historical lottery spend and 2026 forecast"
#| message: false
#| warning: false

historical <- georgia_lottery |>
  group_by(year) |>
  summarise(spend = sum(lottery_spend, na.rm = TRUE) / 1e6, .groups = "drop") |>
  mutate(type = "Historical")

forecast_point <- tibble(
  year = 2026,
  spend = total_predicted / 1e6,
  type = "Forecast"
)

trend_data <- bind_rows(historical, forecast_point)

ggplot(trend_data, aes(x = year, y = spend)) +
  geom_smooth(
    data = historical,
    method = "lm", se = TRUE, color = "#2166ac", fill = "#c6dbef", linewidth = 1
  ) +
  geom_point(aes(color = type, shape = type), size = 4) +
  geom_text(
    data = forecast_point,
    aes(label = paste0("$", round(spend, 1), "M")),
    vjust = -1, fontface = "bold", color = "#d6604d", size = 4
  ) +
  scale_color_manual(values = c("Historical" = "#2166ac", "Forecast" = "#d6604d")) +
  scale_shape_manual(values = c("Historical" = 16, "Forecast" = 18)) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M")) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  labs(
    title = "Total Georgia lottery spend: historical and 2026 forecast",
    x = "Year",
    y = "Total spend (millions)",
    color = NULL,
    shape = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

The sections below break down where that revenue is expected to come from.

---

## Who is buying tickets?

::: {layout-ncol=2}

```{r}
#| label: donut-sex
#| fig-cap: "Predicted 2026 lottery spend by sex"
#| message: false
#| warning: false

sex_totals <- predictions_2026 |>
  group_by(sex) |>
  summarise(spend = sum(lottery_spend, na.rm = TRUE), .groups = "drop") |>
  mutate(
    pct = spend / sum(spend),
    label = paste0(sex, "\n", scales::percent(pct, accuracy = 0.1))
  )

ggplot(sex_totals, aes(x = 2, y = spend, fill = sex)) +
  geom_col(width = 1, color = "white", linewidth = 0.5) +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 4.5, fontface = "bold", color = "white"
  ) +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = c("#2166ac", "#d6604d")) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Predicted lottery spend by sex")
```

```{r}
#| label: donut-race
#| fig-cap: "Predicted 2026 lottery spend by race/ethnicity"
#| message: false
#| warning: false

race_totals <- predictions_2026 |>
  group_by(race_ethnicity) |>
  summarise(spend = sum(lottery_spend, na.rm = TRUE), .groups = "drop") |>
  mutate(
    pct = spend / sum(spend),
    label = paste0(race_ethnicity, "\n", scales::percent(pct, accuracy = 0.1))
  )

race_colors <- c(
  "#2166ac", "#d6604d", "#4dac26", "#762a83",
  "#f4a582", "#1b7837", "#e08214"
)

ggplot(race_totals, aes(x = 2, y = spend, fill = race_ethnicity)) +
  geom_col(width = 1, color = "white", linewidth = 0.5) +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 3.5, fontface = "bold", color = "white"
  ) +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = race_colors) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Predicted lottery spend by race/ethnicity")
```

:::

---

## Spend by age group and income

::: {layout-ncol=2}

```{r}
#| label: bar-age
#| fig-cap: "Total predicted 2026 lottery spend by age group"
#| message: false
#| warning: false

age_totals <- predictions_2026 |>
  group_by(age_group) |>
  summarise(spend = sum(lottery_spend, na.rm = TRUE), .groups = "drop")

ggplot(age_totals, aes(x = age_group, y = spend / 1e6, fill = age_group)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(
    aes(label = paste0("$", round(spend / 1e6, 1), "M")),
    vjust = -0.4, size = 3.5, fontface = "bold"
  ) +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M"), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Predicted lottery spend by age group",
    x = "Age group",
    y = "Predicted spend (millions)"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank())
```

```{r}
#| label: bar-income
#| fig-cap: "Total predicted 2026 lottery spend by income level"
#| message: false
#| warning: false

income_totals <- predictions_2026 |>
  group_by(income_level) |>
  summarise(spend = sum(lottery_spend, na.rm = TRUE), .groups = "drop")

ggplot(income_totals, aes(x = income_level, y = spend / 1e6, fill = income_level)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(
    aes(label = paste0("$", round(spend / 1e6, 1), "M")),
    vjust = -0.4, size = 3.5, fontface = "bold"
  ) +
  scale_fill_brewer(palette = "Oranges", direction = 1) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M"), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Predicted lottery spend by income level",
    x = "Income level",
    y = "Predicted spend (millions)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

:::

---

## Where in Georgia are tickets being sold?

The map below shows predicted per-capita lottery spend for each county in 2026. Hover over a county to see the estimated dollar amount.

```{r}
#| label: leaflet-map
#| message: false
#| warning: false
create_lottery_map(predictions_2026)
```
